Self-Supervised Transfer Learning From 2D Images to 3D Point Clouds

### 2022-05-06 doc/paper.tex
adds semantic segmentation results section

In some experiments, we vary the amount of training labeled data on downstream tasks to see the effectiveness of dataset size on the self-supervised pre-training.

@@

\begin{figure}
    \centering
    \includegraphics[width=7cm]{images/experiments/25.02.2022-image-pretrain-vis2.png}
    \includegraphics[width=7cm]{images/experiments/25.02.2022-image-pretrain-vis1.png}
    \caption{\textbf{Left} Input image. \textbf{Right} Visualization of features from the pre-trained 2D backbone}
    \label{fig:features2dvis}
\end{figure}

@@

We first test the case of pre-training our model on ScannetV2 and fine tuning on S3DIS. Detailed results are shown in table \ref{table:s3disResults}. Fully supervised pre-training has a drastic impact on the final downstream performance ($+5.1$ mIOU) and is used as a pseudo upper bound on expected performance.

Almost all base algorithms improve downstream performance, except DepthContrast. Our algorithm is comparable to PointContrast, both achieving a 1 mIOU increase  but perform worse than CSC.

Table \ref{table:scannetVaryingDataAmount} looks at using weights pre-trained on the entire Scannet dataset but then fine-tuned using only a portion of the labels. This simulates the use-case of capturing a large amount of data and only labeling a subset. All methods show no improvement when given access to labels for the entire dataset. The performance gains are unpredictable, with

Interestingly, tightening the minimum overlap ratio between scan pairs improved performance on S3DIS.

Our method has the advantage of not requiring the transformation between adjacent scans, and so to test how PointContrast and CSC perform under the same restriction we used the same scan

\begin{itemize}
    \item Comparable to point contrast
    \item CSC does best with tight overlap
    \item DepthContrast has no improvement
    \item Point Contrast has no improvement when using the same scan x
\end{itemize}

\begin{table}
    \centering
    \resizebox{0.45\textwidth}{!}{
        \begin{tabular}{ c | c c c c }
                                                                & Scan & Overlap  & Semantic (mIOU)               & Instance (mAP@0.5)            \\
            \hline
            Scratch                                             & --   & --       & 65.1                          & 53.0                          \\
            Supervised                                          & --   & --       & \textbf{70.2} \textbf{(+5.1)} & \textbf{56.2} \textbf{(+3.2)} \\
            \hline
            $\dagger$ PointContrast \cite{xie2020pointcontrast} & Pair & $>$ 30\% & 66.2 \textbf{(+1.1)}          & 54.8 \textbf{(+1.8)}          \\
            PointContrast \cite{xie2020pointcontrast}           & Pair & $>$ 70\% & 68.6 \textbf{(+3.5)}          & 55.3 \textbf{(+2.3)}          \\
            PointContrast  \cite{xie2020pointcontrast}          & Same & --       & 65.0 (-0.1)                   & 56.6 \textbf{(+3.6)}          \\
            $\dagger$ CSC \cite{hou2021Exploring}               & Pair & $>$ 30\% & 69.0 \textbf{(+3.9)}          & 57.8 \textbf{(+4.8)}          \\
            CSC \cite{hou2021Exploring}                         & Pair & $>$ 70\% & \textbf{70.1} \textbf{(+5.0)} & \textbf{58.9} \textbf{(+5.9)} \\
            CSC \cite{hou2021Exploring}                         & Same & --       & 67.9 \textbf{(+2.8)}          & 57.2 \textbf{(+4.2)}          \\
            $\dagger$ DepthContrast \cite{zhang2021Self}        & Same & --       & 64.9 (-0.2)                   & 52.3 (-0.7)                   \\
            \hline
            $\dagger$ CME \textbf{(Ours)}                       & Same & --       & 66.5 \textbf{(+1.4)}          & 55.8 \textbf{(+2.8)}          \\
        \end{tabular}
    }
    \caption{\textbf{S3DIS downstream performance.} All methods use weights pre-trained on Scannet. Those with a $\dagger$ indicate the original algorithm whereas all others are variants. }
    \label{table:s3disResults}
\end{table}

\begin{table}
    \centering
    \resizebox{0.47\textwidth}{!}{
        \begin{tabular}{ c | c c c c c c }
                                                      & 5\%                  & 10\%                 & 20\%                  & 30\%                 & 40\%                 & 100\%                \\
            \hline
            Scratch                                   & 39.2                 & 40.4                 & 41.1                  & 41.7                 & 40.8                 & 41.0                 \\
            \hline
            PointContrast \cite{xie2020pointcontrast} & 40.3 \textbf{(1.1)}  & 42.6 \textbf{(+2.2)} & 42.2  \textbf{(+1.1)} & 41.7 \textbf{(+0.0)} & 43.0 \textbf{(+2.2)} & 42.1 \textbf{(+1.1)} \\
            CSC \cite{hou2021Exploring}               & 40.6 \textbf{(+1.4)} & 42.2 \textbf{(+1.8)} & 42.3 \textbf{(+1.2)}  & 42.2 \textbf{(+0.6)} & 44.3 \textbf{(+3.5)} & 43.0 \textbf{(+2.0)} \\
            DepthContrast \cite{zhang2021Self}        & 39.7 \textbf{(+0.5)} & 40.9 \textbf{(+0.5)} & 41.2 \textbf{(+0.1)}  & 41.9 \textbf{(+0.2)} & --                   & --                   \\
            \hline
            CME \textbf{(Ours)}                       & 39.5 \textbf{(+0.3)} & 40.9 \textbf{(+0.5)} & 40.9 (-0.2)           & 41.8 \textbf{(+0.1)} & 41.8 \textbf{(+1.0)} & 42.0 \textbf{(+1.0)} \\
        \end{tabular}
    }
    \caption{\textbf{KITTI Semantic Segmentation results (mIOU)}. All methods use weights pre-trained on Scannet.  }
    \label{table:kittiSemanticResults}
\end{table}

@@

\begin{figure}
    \centering
    \includegraphics[width=8cm]{images/plots/s3dis_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_0.05_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_semantic.pdf}
    \caption{Comparison of validation performance over training steps for models pre-trained on Scannet and fine-tuned for semantic segmentation on, \textbf{Top:} S3DIS, \textbf{Middle:} Scannet with only 5\% labeled data, \textbf{Bottom:} Scannet}
    \label{fig:validationComparison}
\end{figure}

@@

@@

>>>>

As is common in self-supervised learning experiments, we investigate the effect of varying the amount of labeled data to see the impact pre-training on a large unlabeled dataset has on smaller labeled datasets.

@@

\begin{figure}
    \centering
    \includegraphics[width=6.5cm]{images/experiments/25.02.2022-image-pretrain-vis2.png}
    \includegraphics[width=6.5cm]{images/experiments/25.02.2022-image-pretrain-vis1.png}
    \caption{\textbf{Left} Input image. \textbf{Right} Visualization of features from the pre-trained 2D backbone}
    \label{fig:features2dvis}
\end{figure}

@@

We first test the case of pre-training our model on ScannetV2 and fine-tuning on S3DIS. Detailed results are shown in table \ref{table:s3disResults}. Fully supervised pre-training has a drastic impact on the final downstream performance ($+5.1$ mIOU) and is used as a pseudo upper bound on expected performance. Almost all base algorithms improve downstream results, except DepthContrast. Our algorithm is comparable to PointContrast, both achieving a 1\% mIOU increase but perform worse than CSC.

To simulate the use-case of capturing a large amount of un-labeled data for pre-training and only labeling a subset of it, we took Scannet and partitioned it into different portions of labeled data. The results can be found in table \ref{table:scannetVaryingDataAmount}. All methods show no improvement when given access to labels for the entire dataset. However, as the labeled data is restricted, the pre-trained models start to have an impact, with all methods improving upon training from scratch when given access to only 5\% of the labels.

Similarly, we also test varying data amounts on the semanticKITTI dataset which can be found in table \ref{table:kittiSemanticResults}. We pre-trained all methods on Scannet to see how well the learned representations transferred to different data domains. In almost all cases there are performance improvements with CSC achieving the biggest increases. Unlike in Scannet, all algorithms have some impact across all labeled data ratios. Also having more labeled data increased the performance boost of the pre-trained model, which is the opposite of what we saw in Scannet and S3DIS. This suggests that pre-training on a dataset which is out-of-distribution from the labeled dataset is more informative than pre-training on in-distribution.

\begin{table}
    \centering
    \resizebox{0.45\textwidth}{!}{
        \begin{tabular}{ c | c c c c }
                                                                & Scan & Overlap  & Semantic (mIOU)               & Instance (mAP@0.5)            \\
            \hline
            Scratch                                             & --   & --       & 65.1                          & 53.0                          \\
            Supervised                                          & --   & --       & \textbf{70.2} \textbf{(+5.1)} & \textbf{56.2} \textbf{(+3.2)} \\
            \hline
            $\dagger$ PointContrast \cite{xie2020pointcontrast} & Pair & $>$ 30\% & 66.2 \textbf{(+1.1)}          & 54.8 \textbf{(+1.8)}          \\
            PointContrast \cite{xie2020pointcontrast}           & Pair & $>$ 70\% & 68.6 \textbf{(+3.5)}          & 55.3 \textbf{(+2.3)}          \\
            PointContrast  \cite{xie2020pointcontrast}          & Same & --       & 65.0 (-0.1)                   & 56.6 \textbf{(+3.6)}          \\
            $\dagger$ CSC \cite{hou2021Exploring}               & Pair & $>$ 30\% & 69.0 \textbf{(+3.9)}          & 57.8 \textbf{(+4.8)}          \\
            CSC \cite{hou2021Exploring}                         & Pair & $>$ 70\% & \textbf{70.1} \textbf{(+5.0)} & \textbf{58.9} \textbf{(+5.9)} \\
            CSC \cite{hou2021Exploring}                         & Same & --       & 67.9 \textbf{(+2.8)}          & 57.2 \textbf{(+4.2)}          \\
            $\dagger$ DepthContrast \cite{zhang2021Self}        & Same & --       & 64.9 (-0.2)                   & 52.3 (-0.7)                   \\
            \hline
            $\dagger$ CME \textbf{(Ours)}                       & Same & --       & 66.5 \textbf{(+1.4)}          & 55.8 \textbf{(+2.8)}          \\
        \end{tabular}
    }
    \caption{\textbf{S3DIS downstream performance.} All methods use weights pre-trained on Scannet. Those with a $\dagger$ indicate the original algorithm whereas all others are variants. The difference in brackets is with respect to training from scratch.}
    \label{table:s3disResults}
\end{table}

\begin{table}
    \centering
    \resizebox{0.47\textwidth}{!}{
        \begin{tabular}{ c | c c c c c c }
                                                      & 5\%                  & 10\%                 & 20\%                  & 30\%                 & 40\%                 & 100\%                \\
            \hline
            Scratch                                   & 39.2                 & 40.4                 & 41.1                  & 41.7                 & 40.8                 & 41.0                 \\
            \hline
            PointContrast \cite{xie2020pointcontrast} & 40.3 \textbf{(1.1)}  & 42.6 \textbf{(+2.2)} & 42.2  \textbf{(+1.1)} & 41.7 \textbf{(+0.0)} & 43.0 \textbf{(+2.2)} & 42.1 \textbf{(+1.1)} \\
            CSC \cite{hou2021Exploring}               & 40.6 \textbf{(+1.4)} & 42.2 \textbf{(+1.8)} & 42.3 \textbf{(+1.2)}  & 42.2 \textbf{(+0.6)} & 44.3 \textbf{(+3.5)} & 43.0 \textbf{(+2.0)} \\
            DepthContrast \cite{zhang2021Self}        & 39.7 \textbf{(+0.5)} & 40.9 \textbf{(+0.5)} & 41.2 \textbf{(+0.1)}  & 41.9 \textbf{(+0.2)} & --                   & --                   \\
            \hline
            CME \textbf{(Ours)}                       & 39.5 \textbf{(+0.3)} & 40.9 \textbf{(+0.5)} & 40.9 (-0.2)           & 41.8 \textbf{(+0.1)} & 41.8 \textbf{(+1.0)} & 42.0 \textbf{(+1.0)} \\
        \end{tabular}
    }
    \caption{\textbf{KITTI Semantic Segmentation results (mIOU)}. All methods use weights pre-trained on Scannet. The difference in brackets is with respect to training from scratch.}
    \label{table:kittiSemanticResults}
\end{table}

@@

@@

Our algorithm is the only one that shows a consistent increase across varying Scannet data portions as well as S3DIS. Unlike PointContrast and CSC, our algorithm does not require pose transformations between adjacent scans making it easier to implement.

The performance gains are unpredictable as the fraction of labeled data is varied. A performance boost of 3\% at 30\% labeled data can turn into a performance drop at 20\% labeled data. At the extreme of using 5\% labeled data, all algorithms have a performance boost. This further reinforces the idea that pre-training is most useful when the amount of labeled data is extremely limited. Interestingly, DepthContrast which was the worst performing on S3DIS is the best performing by a large margin on Scannet at 5\% labeled data, again showing the unpredictability of the results.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=8cm]{images/plots/s3dis_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_0.05_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_semantic.pdf}
    \caption{Comparison of validation performance over training steps for models pre-trained on Scannet and fine-tuned for semantic segmentation on: \textbf{Top:} S3DIS, \textbf{Middle:} Scannet with only 5\% labeled data, \textbf{Bottom:} Scannet}
    \label{fig:validationComparison}
\end{figure}
